# Introduction
- In this project, the 2009 ASA Statistical Computing and Graphics Dataset is used.
-  Because the dataset is very large (contains the year between 1987 and 2008), 
- I choose 2006, 2007 data as sub-dataset for the tasks.
- This project mainly focus on 5 problems
    1. When is the best time of day, day of the week, and time of year to fly to minimise delays?
    2. Do older planes suffer more delays?
    3. How does the number of people flying between different locations change over time?
    4. Can you detect cascading failures as delays in one airport create delays in others?
    5. Use the available variables to construct a model that predicts delays.
- Library introduction
    - pandas: process csv data
    - numpy: process matrix
    - matplotlib: data visualization
    - sklearn: training prediction model
# Data wrangling
```{r}
library("dplyr")
library("readr")
library("ggplot2")
```
- Load flight, plane, airport data

```{r}
flight_2007 = read_csv("./dataverse_files/2007.csv")
flight_2006 = read_csv("./dataverse_files/2006.csv")
flight = rbind(flight_2006, flight_2007)
```
```{r}
airport = read_csv("./dataverse_files/airports.csv")
```
```{r}
plane = read_csv("./dataverse_files/plane-data.csv")
```

# Question 1
**When is the best time of day, day of the week, and time of year to fly to minimise delays?**
- Join the flight data with the plain data on tailnum.
- Get the hour of the day by dividing CRSDepTime with 100. Because the origin format is HHSS.
- Group the delay according to 'day of month', 'day of week' and 'hour of day'. Calculate the delay mean, median and average.
- Plot the delay and find the best time.
- Use line plot and box plot to display the median and mean delay of different time.
- We mainly focus on the median delay of different time.

```{r}
flight['CRSDepHourOfDay'] = (flight['CRSDepTime'] %/% 100)
flight['DelayMin'] = flight['DepDelay'] / 60
```

```{r}
dom_delay <- flight %>%
  group_by(DayofMonth) %>%
  summarize(depdelay_mean = mean(DepDelay, na.rm=TRUE), depdelay_median = median(DepDelay, na.rm=TRUE))
dow_delay <- flight %>%
  group_by(DayOfWeek) %>%
  summarize(depdelay_mean = mean(DepDelay, na.rm=TRUE), depdelay_median = median(DepDelay, na.rm=TRUE))
hod_delay <- flight %>%
  group_by(CRSDepHourOfDay) %>%
  summarize(depdelay_mean = mean(DepDelay, na.rm=TRUE), depdelay_median = median(DepDelay, na.rm=TRUE))
```

```{r}
ggplot(dom_delay) +
  geom_line(aes(x = DayofMonth, y = depdelay_mean, color='mean')) +
  geom_line(aes(x = DayofMonth, y = depdelay_median, color='median'))
```

```{r}
ggplot(dow_delay) +
  geom_line(aes(x = DayOfWeek, y = depdelay_mean, color='mean')) +
  geom_line(aes(x = DayOfWeek, y = depdelay_median, color='median'))
```

```{r}
ggplot(hod_delay) +
  geom_line(aes(x = CRSDepHourOfDay, y = depdelay_mean, color='mean')) +
  geom_line(aes(x = CRSDepHourOfDay, y = depdelay_median, color='median'))
```

- As shown in the figure, we mainly focus on the median dep_delay group by time.
- Best time of day is: 5 o'clock, because 5 o'clock has least maximum delay value and lower average delay.
- Best day of week is: 2 because week 2 wih lower median and average delay.
- Best day of month is: 6, 8, 9. All of them has least maximum delay value and lower average and median delay. But the difference between them cannot be shown from the box plot.

# Question2
**Do older planes suffer more delays?**
- Import the plane data and drop NA values
- Join the plane data and the flight data on 'tailnum' column
- Group the delay according to the plane's 'year' column, calculate the delay's mean, average and median.
- Plot the delay and the year with line plot, find if there exists a relationship that older planes suffer more delays.

```{r}
plane = read_csv('./dataverse_files/plane-data.csv')

flight_plane = merge(x=flight, y=plane, by.x='TailNum', by.y='tailnum')
```
```{r}
plane_delay <- flight_plane %>%
  group_by(year) %>%
    summarize(depdelay_mean = mean(DepDelay, na.rm=TRUE), depdelay_median = median(DepDelay, na.rm=TRUE))
```

```{r}
ggplot(plane_delay) + 
  geom_line(aes(x=year, y=depdelay_mean, group=1, color='Mean')) +
  geom_point(aes(x=year, y=depdelay_mean, group=1)) +
  geom_line(aes(x=year, y=depdelay_median, group=1, color='Median')) +
  geom_point(aes(x=year, y=depdelay_median, group=1)) + 
  scale_x_discrete(guide = guide_axis(angle = 90))

```
- As shown in the figure, we mainly focus on the mean dep_delay group by plane's issue year.
- There are no clearly relation between manufacturing year and mean dep_delay time. So older planes do not suffer more delays.

# Question3
**How does the number of people flying between different locations change over time?**
    Find the locations with top-5 flight number and find the change of the number of people according to months (line plot).
    - First combine the origin and destination with the same order, for example, treat (BOS,LGA) and (LGA, BOS) as the same key.
    - Group the data with the origin dest tuple and sort with the count.
    - Get the top-5 origin dest tuple
    - Plot the change of those locations and find the relationship 
    
```{r}
flight <- flight %>%
  mutate(SortedOriginDest = case_when(
    Origin > Dest ~ paste(Dest, ',', Origin),
    Origin <= Dest ~ paste(Origin, ',',Dest)
    ))

N = 5
max_n = flight %>%
  group_by(SortedOriginDest) %>%
  summarize(cnt = n())
max_n = max_n[order(-max_n$cnt),]
max_n = head(max_n, N)
max_n_OD = max_n['SortedOriginDest']
```

```{r}
flight_od <- filter(flight, SortedOriginDest %in% max_n_OD$SortedOriginDest)
flight_od <- flight_od %>% 
  group_by(SortedOriginDest, Month) %>%
  summarize(cnt = n())
```

```{r}
ggplot(flight_od) + 
  geom_line(aes(x=Month, y=cnt, group=SortedOriginDest, color=SortedOriginDest)) +
  geom_point(aes(x=Month, y=cnt, group=SortedOriginDest, color=SortedOriginDest))
```
- In this approach, I have choose the top-5 flight, group by both direction of the Origin and Destination city.
- For example, flight with direction (HNL, OGG) and (OGG, HNL) are count as the same 'locations'.
- As shown in the figure, for the flight with dest or origin of HNL, Feburary has least people flying and July has most people flying.
- All location has a decrease in February and an increase in March.
- People flying in summer is more than people flying in winter.
# Question4
**Can you detect cascading failures as delays in one airport create delays in others?**
- Group the flight data by the time and destination attribute. 
- Aggegrate the mean and median value of the departure delay attribute.
- Join the flight according to the destination and origin attribute within the same day.
    - For example, one flight from LA to DC will join with another record from DC to another place on the same day.
- Because the data is too large, I only choose the data in 2006, Jan
- Calculate the ratio between the destination and origin's departure delay, if the ratio is steady and close to 1.
- Use the scatter plot to plot the ratio and time in different airports.

```{r}
flight_delay <- flight %>% filter(DepDelay > 0)
flight_delay <- flight_delay %>% filter(Year==2006)
flight_delay <- flight_delay %>% filter(Month==1)
flight_sub_dest_grouped = flight_delay %>% 
  group_by(Year, Month, DayofMonth, Dest) %>%
  summarize(depdelay_mean = mean(DepDelay, na.rm=TRUE), depdelay_median = median(DepDelay, na.rm=TRUE))
flight_sub_orig_grouped = flight_delay %>% 
  group_by(Year, Month, DayofMonth, Origin) %>%
  summarize(depdelay_mean = mean(DepDelay, na.rm=TRUE), depdelay_median = median(DepDelay, na.rm=TRUE))
flight_sub_joined = merge(x=flight_sub_dest_grouped, y=flight_sub_orig_grouped, 
                          by.x=c("Year", 'Month', 'DayofMonth', 'Dest'),
                          by.y=c("Year", 'Month', 'DayofMonth', 'Origin'))
```
```{r}
flight_sub_joined['ratio'] = flight_sub_joined['depdelay_median.x'] / flight_sub_joined['depdelay_median.y']
```

```{r}
flight_sub_joined <- flight_sub_joined %>% filter(ratio < 2)
ggplot(flight_sub_joined, aes(x = DayofMonth, y = ratio)) +
  geom_point()
```

- As shown in the figure, the ratio of the origin and destination delay does not have clearly relationship, the ratio is not close to 1.
- So there is no cascading failures as delays in one airport create delays in others.

# Question5
**Use the available variables to construct a model that predicts delays.**
- Divide the dataset into two groups: delay and no delay
- Join the flight data with the airport, with both origin and destination airport info.
- Choose time, distance and location as attribute
    - "Year", "Month", "DayOfWeek", "CRSDepTime", "CRSArrTime", "CRSElapsedTime", "Distance", "lat_x", "long_x", "lat_y", "long_y"
- Downsample the data with 20000 items, because the origin dataset is too large.
- Divide the dataset into train and test with 7 : 3 ratio.
- Train the model of random forest
- Get the test result

```{r}
airport = read_csv('./dataverse_files/airports.csv')
sampled_flight = flight[sample(nrow(flight), 20000), ]
flight_ori_airport = merge(sampled_flight, airport, by.x='Origin', by.y='iata')
flight_src_airport = merge(flight_ori_airport, airport, by.x='Dest', by.y='iata')
flight_src_airport['delay'] = flight_src_airport['ActualElapsedTime'] - flight_src_airport['CRSElapsedTime']
```
```{r}
flightds =  flight_src_airport %>% select("Year", "Month", "DayOfWeek", "CRSDepTime", "CRSArrTime", "CRSElapsedTime", "Distance", 'delay', "lat.x", "long.x", "lat.y", "long.y")
flightds <- flightds %>%
  mutate(is_delay = case_when(
    delay > 0 ~ 1,
    delay <= 0 ~ 0
    ))
```
```{r}
library(randomForest)
```